---
title: "Coursera/Reproducible Research Project 1"
author: "Richard Ferro"
date: "12/28/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Reading the data into activity data frame.
```{r}
activity <- read.csv("activity.csv", na.strings="NA")
head(activity)
```

Process/transform the data into a format suitable analysis.
```{r}
activity$date <- as.Date(activity$date)
act <- subset(activity, !is.na(activity$steps))
```

## The mean total number of steps taken per day.

```{r}
dailysum <- tapply(act$steps, act$date, sum, na.rm=TRUE, simplify=TRUE)
dailysum <- dailysum[!is.na(dailysum)]

hist(x=dailysum,
     col="blue",
     breaks=25,
     xlab="Daily total steps",
     ylab="Frequency",
     main="The distribution of daily total (missing data ignored)")
     
rmean <- mean(dailysum)
rmedian <- median(dailysum)
```
The mean is `r rmean` and the median is `r rmedian`

## The average daily activity pattern.
 Calculate average steps for each interval for all days. 
 Plot the Average Number Steps per Day by Interval. 
 Find interval with most average steps. 
 
```{r}
a <- tapply(act$steps, act$interval, mean, na.rm=TRUE, simplify=TRUE)
ActivityAvg <- data.frame(interval=as.integer(names(a)), avg=a)

with(ActivityAvg,
     plot(interval,
          avg,
          type="l",
          xlab="5-minute intervals",
          ylab="average steps in the interval across all days"))
 max_steps <- max(ActivityAvg$avg)
 ival <- ActivityAvg[ActivityAvg$avg == max_steps, ]
```

 The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r ival`.
 
## Impute missing values. Compare imputed to non-imputed data.
Report the total number of missing values in the dataset
(i.e. the total number of rows with NAs) 
 
```{r}
sm <- sum(is.na(activity$steps))
```
The original data set has `r sm` rows with missing data.
A simple strategy for filling in all of the missing values in the dataset.
If a 5-minute interval has missing value, then the mean for that 5-minute interval.
A new data frame is created, activity1, that is equal to the original dataset but with the missing data filled in (using mean for that interval for imputation).

```{r}
activity1 <- activity
nx <- is.na(activity1$steps)
a <- tapply(act$steps, act$interval, mean, na.rm=TRUE, simplify=TRUE)
activity1$steps[nx] <- a[as.character(activity1$interval[nx])]
```

Make a histogram of the total number of steps taken each day.
```{r}
new_dailysum <- tapply(activity1$steps, activity1$date, sum, na.rm=TRUE, 
                       simplify=TRUE)
hist(x=new_dailysum,
     col="blue",
     breaks=20,
     xlab="daily steps",
     ylab="frequency",
     main="The distribution of daily total (with missing data imputed)")
```

Calculate and report the mean and median total number of steps taken per day.

```{r}
imp_mean <- mean(new_dailysum)
imp_median <- median(new_dailysum)
```
The imputed mean is `r imp_mean` and the imputed median is `r imp_median`

Calculate difference between imputed and non-imputed data.
```{r}
mean_diff <- imp_mean - rmean
med_diff <- imp_median - rmedian
```
The difference in the means `r mean_diff`.
The difference in the medians `r med_diff`.

## Differences in activity patterns between weekdays and weekends.

```{r}
is_weekday <- function(d) {
  wd <- weekdays(d)
  ifelse (wd == "Saturday" | wd == "Sunday", "weekend", "weekday")
}

wx <- sapply(activity1$date, is_weekday)
activity1$wk <- as.factor(wx)
head(activity1)

wk_act <- aggregate(steps ~ wk+interval, data=activity1, FUN=mean)

library(lattice)

xyplot(steps ~ interval | factor(wk),
       layout = c(1, 2),
       xlab="Interval",
       ylab="Number of steps",
       type="l",
       lty=1,
       data=wk_act)
```


